### 2023-12-26 pnrs.tex
advancing in the results

@@

@@

\begin{theorem}
    The \kl{$\Nat$-rational polynomials} are exactly
    the polynomials in $\CorrectPoly$.
\end{theorem}

@@

decide nat inside rel in the commutative input case.

@@

>>>>

Let us assert that polynomials that have all coefficients in $\Nat$ are
\kl{$\Nat$-rational polynomials}, which will be a base case in an upcoming
induction.

@@

\begin{lemma}
    Let $f$ be a  \kl{$\Nat$-polyregular} function. 
    There exists $\omega \in \Nat$
    such that for all $p \in \Nat$,
    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
    for all $u_1, \dots, u_p \in \Sigma^*$,
    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
    whose \kl{maximal monomials} have \kl{positive coefficients},
    and such that:
    \begin{equation*}
        f(\alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i))
        = P(X_1, \dots, X_p) \quad .
    \end{equation*}
\end{lemma}

@@

\begin{theorem}
    \label{corrected-version:thm}
    The \kl{$\Nat$-rational polynomials} are exactly
    the polynomials in $\CorrectPoly$.
\end{theorem}

@@

The goal of this section is to go beyond polynomials and characterize all
$\Rel$-polyregular functions that are $\Nat$-polyregular, under the assumption
that they are \kl{commutative}.

\begin{theorem}
    \label{decidable-n-poly:thm}
    Given a \kl{commutative}
    \kl{$\Rel$-polyregular function} $f \colon \Sigma^* \to \Rel$,
    one can decide if $f$ can be computed by a \kl{$\Nat$-polyregular function},
    and effectively compute a representation of $f$.
\end{theorem}

\begin{lemma}
    \label{decide-correct-poly:lem}
    One can decide if a polynomial is in $\CorrectPoly$.
\end{lemma}
\begin{proof}
    It is easy to check that maximal monomials have positive coefficients.
    Assuming that it is the case, remark that 
    every partial valuation $\nu$
    such that $\restr{P}{\nu}$ has a negative maximal coefficient
    must have values that are bounded by the sum of absolute values of
    coefficients appearing in $P$.
    In particular, one can check all possible choices for $\nu$
    within these bounds.
\end{proof}

To prove the theorem, let us first provide a nice characterization of
\kl{commutative} \kl{$\Rel$-polyregular functions}
in terms of polynomials.

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function}. There exists $\omega \in \Nat$,
    and polynomials $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{ 0 \leq n_a < \omega}$,
    such that for all $w \in \Sigma^*$,
    \begin{equation*}
        f(w) = P_{\seqof{\card[a]{w} [\omega]}{a \in \Sigma}}
        (\seqof{\card[a]{w} / \omega}{a \in \Sigma}) \quad .
    \end{equation*}
    Which informally means that for every choice of rest modulo $\omega$
    for every variable, there exists a polynomial describing the
    behavior of $f$ when adding multiples of $\omega$ letters at a time.

    Furthermore, the polynomials are computable from $f$.
\end{lemma}
\begin{proof}
\end{proof}

As a consequence, we conclude 
\begin{proof}[Proof of \cref{decidable-n-poly:thm}]
    We use \cref{decompose-polynomial:lem}
    to effectively compute
    a polynomial representation of $f$. Remark that
    $f$ is a \kl{$\Nat$-polyregular function} if and only if
    all these polynomials are \kl{$\Nat$-rational polynomials},
    which is decidable thanks to
    \cref{decide-correct-poly:lem,corrected-version:thm}.
\end{proof}

@@

\begin{conjecture}
    $\CorrectPoly = \setof{P}{ P \geq 0 \wedge \exists K \in \Nat, P \circ (+K) \in \Nat[\vec{X}]}$.
\end{conjecture}
\begin{proof}[Proof Sketch]
    If $S \defined P \circ (+K) \in \Nat[\vec{X}]$, then for every partial application,
    maximal coefficients of $S$ are positive. As maximal coefficients of 
    (partial applications of) $S$ are the same as the ones of $P$
    \emph{except for the constant terms} (full application),
    where $P$ is already guaranteed to be non-negative, we conclude
    that $P \in \CorrectPoly$.

    Conversely, we prove by induction on the degree of $P \in \CorrectPoly$,
    that there exists some $K$ such that $P \circ (+K) \in \Nat[\vec{X}]$.
    First, notice that $\CorrectPoly$ is closed under pre-composition with
    $(+K)$. Then, remark that $P(X - K)$ has the same maximal coefficients as
    $P$ and new maximal coefficients that are multiples of $K$. Hence, for a
    large enough $K$, we conclude that $P(X - K)$ is the sum of its maximal
    coefficients, plus a polynomial that lies in $\CorrectPoly$. We conclude by
    induction hypothesis. 
\end{proof}

\begin{conjecture}
    $\CorrectPoly$ is precisely composed of polynomials that
    have non-negative coefficients in a \emph{binomial basis},
    i.e., where polynomials are obtained
    as sums and products of $\binom{X_i}{p_i}$.
\end{conjecture}
